Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-01-28 joss/paper.Rmd
Kindle edits

The growth of programming in the sciences has been explosive in the
last decade. In particular, the statistical programming language ``R``
has grown exponentially to become one of the top ten programming
languages in use today. Increasingly, concerns have arisen over the
reproducibility of scientific research [@Peng2011 @Baker2016
@Stodden2018] and the potential issues stemming from the complexity
and fragility of analytical software [@Pasquier2017]. There is now a
recognition that simply making the code open is not enough, and that
there is a need for improvements to documentation and transparency
[@Pasquier2018].

At it's root R is a statistical programming language, that is, it was
designed for use in analytical workflows. As such, the majority of the
R community is focused on producing code for idiosyncratic projects
that are results oriented. Also, R's design is intentionally at a
level that abstracts many aspects of programming that would otherwise
act as a barrier to entry for many users. This is good in that there
are many people who use R to their benefit with little to no formal
training in computer science or software engineering. However, these
same users are also frequently frustrated by code that is fragile,
buggy and complicated enough to quickly become obtuse (even to
themselves) in a very short amount of time, which leads to frequently
re-writing code unecessarily. Also, when scripts take an extremely
long time to execute, reducing unnecessary analyses can increase
computational efficiency.

The stability, reproducibility and re-use of scientific analyses in R
would be improved by refactoring. From this perspective, tools that
can lower the time and energy required to re-factor and streamline
analytical scripts, and otherwise help to "clean" code will have a
significant impact on scientific reproducibility across all
disciplines [@Visser2015]. More often then not, when someone is
writing an R script, the intent to produce a set of results, such as a
statistical analysis, figure, etc. This set of results is always a
subset of a much larger set of possible ways to explore a dataset, as
there are many statistical approaches and tests, let alone ways to
create visualizations and other representations of patterns in
data. This commonly leads to lengthy, complicated scripts from which
researchers manually subset results, but never refactor (i.e. reduce
to the desired subset). In part, this is driven a desire to record the
entire scripting/analytical process and not lose work, which is
alleviated by the use of proper version control systems (e.g. git and
subversion); however, refactoring will always be a part of the
analytical process regardless.

To support provide support for easier refactoring in R, we have
created ``Rclean``. The ``Rclean`` package provides tools to
automatically reduce a script to the parts that are specifically
relevant to a research product (e.g.  a scientific report, academic
talk, research article, etc.). Although potentially useful to all R
coders, it was designed to ease refactoring for scientists that use R
but do not have formal training in software engineering. Here, we
detail the structure of the package's API, describe the general
workflow illustrated by an example use case and provide some
background on how data provenance enables the underlying functionality
of the package. We then end with a discussion of future applications
of data provenance in the context of "code cleaning" and the potential
integration with other software engineering tools for the R community.

@@

The main goal of *Rclean* is to provide a means to simplify code. To
keep the process simple and straight-forward, the API has been kept to
a minimum set of functions, which enable a user to conduct the basic
workflow of getting information about the posible results in a script,
obtaining the minimum code for a set of results and then creating a
new "cleaned" script or other software (e.g. a function, reproducible
example, web-app, etc.).

@@

In the process of cleaning a script, it is likely that a user might
require some help analyzing the script. There are several functions to
help. The `get_vars` function will return a list of possible results
for a given script at a supplied file path. This is obviously an
important step, and justsifiably, the default behavior of the `clean`
function is to run `get_vars` if no results are supplied. To help with
limiting and checking the selection of results, the `code_graph` function
creates a network graph of the relationships among the various
results and lines of code in the script. Last, the `get_libs` function
can be used to detect the packages that a given script depends on,
which it will return as coded library calls that can be inserted into
a cleaned script.

All of these processes rely on the generation of data provenance
[@Carata2014], which is a formal representation of the execution of a
computational process (https://www.w3.org/TR/prov-dm/), to rigorously
determine the the unique computational pathway from inputs to
results. To avoid confusion, note that "data" in this context is used
in a broad sense to include all of the information generated during
computation, not just the data that are collected in a research
project that are used as input to an analysis. Having the formalized,
mathematically rigorous representation that data provenance provides
gaurantees that the analyses that *RClean* conducts are theoretically
sound. Most importantly, it makes it possible to apply a network
search algorithm to determine the minimum and sufficient code needed
to generate the chosen result in the `clean` function.

There are multiple approaches to collecting data provenance, but
*Rclean* uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. *Rclean* relies on a library called
*CodeDepends* to gather the prospective provenance for each
script. For more information on the mechanics of the *CodeDepends*
package, see [@Lang2019]. To get an idea of what data provenance is,
take a look at the `code_graph` function. The plot that it generates
is a graphical representation of the prospective provenance generated
for *Rclean*.

@@

```{R prov-graph}

@@

Manually tracing through our code for all the variables used in the
test and finding all of the lines that were used to prepare them for
the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. Instead, we can easily do this automatically with
*Rclean*.

@@

As you can see, *Rclean* has picked through the tangled bits of code
and found the minimal set of lines relevant to our object of
interest. This code can now be visually inspected to adapt the
original code or ported to a new, "refactored" script.

>>>>

The growth of programming in the sciences has been explosive in the
last decade. This has facilitated the rapid advancement of scienqce
through the agile development of computational tools. However,
concerns have begun to surface about the reproducibility of scientific
research in general [@Peng2011 @Baker2016] and the potential issues
stemming from issues with analytical software [@Pasquier2017
@Stodden2018]. Specifically, there is a growing recognition across
disciplines that simply making data and software "available" is not
enough and that there is a need to improve the transparency and
stability of scientific software [@Pasquier2018].

At the core of the growth of scentific conputation, the ``R``
statistical programming language has grown exponentially to become one
of the top ten programming languages in use today. At it's root R is a
*statistical* programming language. That is, it was designed for use
in analytical workflows; and the majority of the R community is
focused on producing code for idiosyncratic projects that are
*results* oriented. Also, R's design is intentionally at a level that
abstracts many aspects of programming that would otherwise act as a
barrier to entry for many users. This is good in that there are many
people who use R to their benefit with little to no formal training in
computer science or software engineering, but these same users can
also be frequently frustrated by code that is fragile, buggy and
complicated enough to quickly become obtuse even to the authors. The
stability, reproducibility and re-use of scientific analyses in R
would be improved by refactoring. From this perspective, tools that
can lower the time and energy required to re-factor and streamline
analytical scripts and otherwise help to "clean" code, but abstracted
enough to be easily accessible, could have a significant impact on
scientific reproducibility across all disciplines [@Visser2015].

To provide support for easier refactoring in R, we have created
``Rclean``. The ``Rclean`` package provides tools to automatically
reduce a script to the parts that are specifically relevant to a
research product (e.g.  a scientific report, academic talk, research
article, etc.). Although potentially useful to all R coders, it was
designed to ease refactoring for scientists that use R but do not have
formal training in software engineering. Here, we detail the structure
of the package's API, describe the general workflow illustrated by an
example use case and provide some background on how data provenance
enables the underlying functionality of the package. We then end with
a discussion of future applications of data provenance in the context
of "code cleaning" and the potential integration with other software
engineering tools for the R community.

@@

More often then not, when someone is writing an R script, the intent
to produce a set of results, such as a statistical analysis, figure,
table, etc. This set of results is always a subset of a much larger
set of possible ways to explore a dataset, as there are many
statistical approaches and tests, let alone ways to create
visualizations and other representations of patterns in data. This
commonly leads to lengthy, complicated scripts from which researchers
manually subset results, but never refactor, i.e. refine code so that
it is shorter and focused on a desired product.

The goal of `Rclean` is to provide a set of tools that help someone
reduce and organize code based on results. The package uses an
automated technique based on data provenance (details below) to
analyze existing scripts and provide ways to identify and extract code
to produce a desired output. To keep the process simple and
straight-forward much of this process has been abstracted for the
user, and the API has been kept to a minimum set of functions that
enable a user to conduct the following basic workflow:

1. Obtain the "cleaned" code for a result(s).
2. Transfer the code to a new context (e.g. a new script,
function, reproducible example, web-app, etc.).
3. Get information about the posible results and repeat as needed.

## The API

@@

In the process of cleaning a script, it is likely that a user might
require some help analyzing the script. There are several functions to
help with this process. The `get_vars` function will return a list of
possible results for a given script at a supplied file path. This is
obviously an important step, and justsifiably, the default behavior of
the `clean` function is to run `get_vars` if no results are
supplied. To help with limiting and checking the selection of results,
the `code_graph` function creates a network graph of the relationships
among the various results and lines of code in the script. Last, the
`get_libs` function can be used to detect the packages that a given
script depends on, which it will return as coded library calls that
can be inserted into a cleaned script.

## Data Provenance

All of these processes rely on the generation of data provenance.  The
term provenance means information about the origins of some
object. Data provenanve is a formal representation of the execution of
a computational process (https://www.w3.org/TR/prov-dm/), to
rigorously determine the the unique computational pathway from inputs
to results [@Carata2014]. To avoid confusion, note that "data" in this
context is used in a broad sense to include all of the information
generated during computation, not just the data that are collected in
a research project that are used as input to an analysis. Having the
formalized, mathematically rigorous representation that data
provenance provides gaurantees that the analyses that *RClean*
conducts are theoretically sound. Most importantly, because the
relationships defined by the provenance can be represented as a graph,
it is possible to apply network search algorithms to determine the
minimum and sufficient code needed to generate the chosen result in
the `clean` function.

There are multiple approaches to collecting data provenance, but
`Rclean` uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. `Rclean` relies on a library called
`CodeDepends` to gather the prospective provenance for each
script. For more information on the mechanics of the `CodeDepends`
package, see [@Lang2019]. To get an idea of what data provenance is,
take a look at the `code_graph` function. The plot that it generates
is a graphical representation of the prospective provenance generated
for `Rclean`.

@@

	```{R prov-graph, fig.cap = "Network diagram of the prospective data provenance generated for an example script. Arrows indicate relationships among objects and lines of code (numbered)."}

@@

Manually tracing through our code for all the variables used in the
test and finding all of the lines that were used to prepare them for
the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. Instead, we can easily do this automatically with
`Rclean`.

@@

As you can see, `Rclean` has picked through the tangled bits of code
and found the minimal set of lines relevant to our object of
interest. This code can now be visually inspected to adapt the
original code or ported to a new, "refactored" script.
